Reanalysis Step 4: Main Effects Outcome Analysis

### 2025-08-27 reanalysis-4-outcome_analysis.qmd
fix

library(mice)
library(survey)
library(broom)
library(marginaleffects)
library(emmeans)

@@

if (file.exists(models_cache_file) && !force_refit) {
  # Load cached models
  cached_models <- readRDS(models_cache_file)
  fitted_models <- cached_models$fitted_models
  model_summaries <- cached_models$model_summaries
  cat("Loaded cached models from:", models_cache_file, "\n")
  cat("To refit models, set force_refit <- TRUE\n")
} else {
  # Fit new models
  cat("Fitting models to each imputed dataset...\n")

@@

  for (i in 1:n_imputations) {
    cat("Fitting model", i, "of", n_imputations, "\r")

    # Get dataset and weights
    imp_data <- imputed_datasets[[i]]
    imp_data$weights <- twang_weights[[i]]

@@

  cat("\n")

@@

  saveRDS(cached_models, models_cache_file)
  cat("Models cached to:", models_cache_file, "\n")
}

# Show first model summary as example
cat("Example model summary (Imputation 1):\n")
kable(model_summaries[[1]] |> 
      select(term, estimate, std.error, p.value) |> 
      head(10))
```

@@

if (file.exists(ame_cache_file) && !force_recalc_ame) {
  # Load cached AMEs
  ame_cache <- readRDS(ame_cache_file)
  ame_list <- ame_cache$ame_list
  cat("Loaded cached AMEs from:", ame_cache_file, "\n")
} else {
  # Calculate AME for each imputed dataset then pool
  cat("Calculating AMEs for each imputed dataset...\n")
  ame_list <- list()

  for (i in 1:n_imputations) {
    cat("Calculating AME", i, "of", n_imputations, "\r")
    imp_data <- imputed_datasets[[i]]
    imp_data$weights <- twang_weights[[i]]

@@

  cat("\n")

  # Save to cache
  ame_cache <- list(
    ame_list = ame_list,
    n_imputations = n_imputations,
    timestamp = Sys.time()
  )
  saveRDS(ame_cache, ame_cache_file)
  cat("AMEs cached to:", ame_cache_file, "\n")
}

@@

@@

# Get estimated marginal means for key combinations
emm_results <- tryCatch({
  # Get predictions for lapse x dr_severity x any_treatment
  emmeans(fitted_models[[1]], 
          ~ ever_lapse_binary * dr_severity * any_treatment,
          type = "response",  # Get probabilities not log-odds
          weights = "proportional")  # Use observed proportions for other variables
}, error = function(e) {
  # If emmeans fails, try a simpler specification
  emmeans(fitted_models[[1]], 
          ~ ever_lapse_binary + dr_severity + any_treatment,
          type = "response",
          weights = "proportional")
})

# Convert to data frame for display
if (!is.null(emm_results)) {
  pred_df <- as.data.frame(emm_results)

@@

  # Create lapse and treatment labels
  pred_df$Lapse <- ifelse(pred_df$ever_lapse_binary == 1, "Lapsed", "Not Lapsed")
  pred_df$Treatment <- ifelse(pred_df$any_treatment == 1, "Treated", "Not Treated")

  # Select and reshape for display
  library(tidyr)
  pred_wide <- pred_df |>
    select(Lapse, dr_severity, Treatment, Predicted_Prob) |>
    pivot_wider(names_from = Lapse, values_from = Predicted_Prob) |>
    arrange(dr_severity, Treatment)

  kable(pred_wide, 
        caption = "Predicted probability of VI/blindness by lapse status (from first imputation)",
        col.names = c("DR Severity", "Treatment", "Not Lapsed", "Lapsed"),
        digits = 3)

  # Calculate contrasts for risk differences
  # This properly accounts for the survey design
  contrasts_lapse <- tryCatch({
    pairs(emmeans(fitted_models[[1]], 
                  ~ ever_lapse_binary | dr_severity * any_treatment,
                  type = "response"))
  }, error = function(e) {
    NULL
  })

  if (!is.null(contrasts_lapse)) {
    contrasts_df <- as.data.frame(contrasts_lapse)
    kable(contrasts_df |> 
          select(dr_severity, any_treatment, estimate, SE, p.value) |>
          mutate(
            Treatment = ifelse(any_treatment == 1, "Treated", "Not Treated"),
            Risk_Difference = round(estimate, 3),
            SE = round(SE, 4),
            p.value = format.pval(p.value, digits = 4, eps = 0.0001)
          ) |>
          select(dr_severity, Treatment, Risk_Difference, SE, p.value),
          caption = "Risk differences (Lapsed - Not Lapsed) with proper variance estimation",
          col.names = c("DR Severity", "Treatment", "Risk Diff", "SE", "P-value"))
  }
} else {
  cat("Note: Predicted probabilities using emmeans could not be calculated.\n")
  cat("The main analysis results above remain valid.\n")
}
```

>>>>

library(mice)
library(survey)
library(broom)
library(marginaleffects)
library(emmeans)
library(tidyr)

@@

if (file.exists(models_cache_file) && !force_refit) {
  # Load cached models
  cached_models <- readRDS(models_cache_file)
  fitted_models <- cached_models$fitted_models
  model_summaries <- cached_models$model_summaries
  models_loaded_from_cache <- TRUE
  cache_location <- models_cache_file
} else {
  # Fit new models
  models_loaded_from_cache <- FALSE

@@

  for (i in 1:n_imputations) {
    # Get dataset and weights
    imp_data <- imputed_datasets[[i]]
    imp_data$weights <- twang_weights[[i]]

@@

@@

  saveRDS(cached_models, models_cache_file)
  cache_location <- models_cache_file
}
```

**Models cache status:** `r ifelse(models_loaded_from_cache, paste("Loaded from", cache_location), paste("Newly fitted and saved to", cache_location))`

### Example Model Summary (First Imputation)

```{r model-summary-display}
kable(model_summaries[[1]] |> 
      select(term, estimate, std.error, p.value) |> 
      head(10),
      caption = "First 10 coefficients from imputation 1")
```

@@

if (file.exists(ame_cache_file) && !force_recalc_ame) {
  # Load cached AMEs
  ame_cache <- readRDS(ame_cache_file)
  ame_list <- ame_cache$ame_list
  ame_loaded_from_cache <- TRUE
  ame_cache_location <- ame_cache_file
} else {
  # Calculate AME for each imputed dataset then pool
  ame_loaded_from_cache <- FALSE
  ame_list <- list()

  for (i in 1:n_imputations) {
    imp_data <- imputed_datasets[[i]]
    imp_data$weights <- twang_weights[[i]]

@@

  # Save to cache
  ame_cache <- list(
    ame_list = ame_list,
    n_imputations = n_imputations,
    timestamp = Sys.time()
  )
  saveRDS(ame_cache, ame_cache_file)
  ame_cache_location <- ame_cache_file
}

@@

**AME cache status:** `r ifelse(ame_loaded_from_cache, paste("Loaded from", ame_cache_location), paste("Newly calculated and saved to", ame_cache_location))`

@@

# Set emmeans options to handle larger reference grids
emm_options(rg.limit = 50000)

# Get estimated marginal means for the main effect of lapse
emm_results <- emmeans(fitted_models[[1]], 
                      ~ ever_lapse_binary,
                      type = "response",  # Get probabilities not log-odds
                      weights = "proportional")  # Use observed proportions

# Display the main effect predictions
pred_df <- as.data.frame(emm_results)

@@

# Create lapse labels
pred_df$Lapse_Status <- ifelse(pred_df$ever_lapse_binary == 1, "Lapsed", "Not Lapsed")

# Display predictions
kable(pred_df |> select(Lapse_Status, Predicted_Prob, SE, lower.CL, upper.CL), 
      caption = "Predicted probability of VI/blindness by lapse status (marginalized over all covariates)",
      col.names = c("Lapse Status", "Predicted Probability", "SE", "Lower CI", "Upper CI"),
      digits = 3)

# Calculate and display the contrast (risk difference)
contrasts_lapse <- pairs(emm_results)
contrasts_df <- as.data.frame(contrasts_lapse)

kable(contrasts_df |> 
      mutate(
        Contrast = "Lapsed - Not Lapsed",
        Risk_Difference = round(estimate, 3),
        SE = round(SE, 4),
        CI_Lower = round(lower.CL, 3),
        CI_Upper = round(upper.CL, 3),
        p.value = format.pval(p.value, digits = 4, eps = 0.0001)
      ) |>
      select(Contrast, Risk_Difference, SE, CI_Lower, CI_Upper, p.value),
      caption = "Main effect: Risk difference with proper variance estimation",
      col.names = c("Contrast", "Risk Difference", "SE", "Lower CI", "Upper CI", "P-value"))
```
